---
title: "pp346_ps1_steiner"
author: "erika steiner"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    css: "style.css"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, results = 'hide'}
knitr::opts_chunk$set(echo = FALSE)

# load packages
# haven for reading in stata data with labels (suggested in
# https://martinctc.github.io/blog/working-with-spss-labels-in-r/)
library("haven")
library("tidyverse")
library("magrittr")
library('knitr')
library('kableExtra')
library("broom")
library("stringr")
library("sandwich")
library("lmtest")
library("car")
library("stargazer")

# get rid of scientific notation
options(scipen=999)
```

## Overarching Questions

(A) How did FTP change the behavior of welfare recipients?

(B) What can we say about the effects of the time limit as opposed to
    other components of the program?

## Analysis Questions

(2) *Open the file ftp_ar.dta. What is the number of observations? Of
    observations in the treatment and control groups? Do your numbers
    match those from the documentation?*

```{r import data}
ftp_ar <- read_dta("ftp_ar.dta")
```

There are `r nrow(ftp_ar)` observations in the public use administrative
records file. The treatment and control group observations are as
follows:

```{r obs table}
# create table with number of observations, separated by rx and control
n_obs <- 
  # establish data frame
  data.frame(ftp_ar %>% # base on admin data file
                      # count number of people with treatment dummy (e) = 1 (rx)
                      # and count number of people e = 0 (control)
                      count(e)) %>% 
  # convert "0"/"1" to "control"/"treatment"
  mutate(e = ifelse(e == 1, "Treatment", "Control")) %>% 
  # set n column name for readability
  setNames(c('Group', 'Number of Observations'))

# print the table
kable(n_obs, 
      # add a title
      caption = "Observations by Treatment Status") %>% 
  # set style 
  kable_styling(bootstrap_options = c("striped", "hover"))

# https://stackoverflow.com/questions/34446139/caption-font-color-with-kable 

```

These observation counts match those shared on page 8 of the FTP
Documentation.

(4) *Verify the data by replicating the control-group means in the file
    “Annotated AR table 1_U.xls.” Restrict attention to the following
    subset of variables:*

<!-- -->

(a) *quarterly employment, quarter of RA to quarter 18;*

```{r control qemp mean}
# create vector of employment quarter columns 1-19

# for quarter of RA to quarter 18, this will be quarter 1 to 19 in data set
# see data documentation pg. 19 for more info
empq_cols <- c("empq1", "empq2", "empq3", "empq4", "empq5", "empq6", "empq7", "empq8", "empq9", "empq10", "empq11", "empq12", "empq13", "empq14", "empq15", "empq16", "empq17", "empq18", "empq19")

c_empq_means_table <- 
  # use admin data set
  ftp_ar %>% 
  # filter for control group
  filter(e == 0) %>% 
  # isolate employment quarter columns
  select (all_of(empq_cols)) %>% 
  # calculate the mean of each column
  summarise(across(everything(), ~ mean (., na.rm = TRUE))) %>% 
  # convert means to percentages (multiply by 100)
  mutate(across(everything(), ~ . * 100)) %>% 
  # rename quarters for readability
  setNames(c("Quarter of RA*", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8","Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18" )) %>% 
  # TK https://stackoverflow.com/questions/61935149/transpose-of-data-frame-turns-first-row-into-a-list
  # transpose for readability
  pivot_longer(., cols = everything()) %>% 
  setNames(c("Quarter", "Employment Rate (%)"))

# print table
kable(c_empq_means_table, 
      # add a title
      caption = "Average Quarterly Employment Rates among Control Group",
      # round to tenths place to mimic original table
      digits = 1) %>% 
  # set style
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  # add footnote to define "RA"
  footnote(symbol = "RA = Random Assignment")
```

(b) *any AFDC/TANF payments, quarter of RA to quarter 18;*

```{r control any afdc mean}
any_afdc_cols <- c("recc1", "recc2", "recc3", "recc4", "recc5", "recc6", "recc7", "recc8", "recc9", "recc10", "recc11", "recc12", "recc13", "recc14", "recc15", "recc16", "recc17", "recc18", "recc19")

c_afdc_means_table <-
  # use admin data set
  ftp_ar %>% 
  # filter for control group
  filter(e == 0) %>% 
  # isolate afdc/tanf payment columns
  select (all_of(any_afdc_cols)) %>% 
  # calculate the mean of each column
  summarise(across(everything(), ~ mean (., na.rm = TRUE))) %>% 
  # convert means to percentages (multiply by 100)
  mutate(across(everything(), ~ . * 100)) %>% 
  # rename quarters for readability
  setNames(c("Quarter of RA*", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8", "Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18")) %>% 
  # transpose for readability
  pivot_longer(., cols = everything()) %>% 
  setNames(c("Quarter", "Any AFDC/TANF Receipt (%)"))

# print table
kable(c_afdc_means_table, 
      # add a title
      caption = "Average Quarterly AFDC/TANF Receipt Rates among Control Group",
      # round to tenths place to mimic original table
      digits = 1) %>% 
  # set style
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  # add footnote to define "RA"
  footnote(symbol = "RA = Random Assignment")


```

(b) *any Food Stamp payments, quarter of RA to quarter 18;*

```{r control any food stamp mean}
rfsc_cols <- c("rfsc1", "rfsc2", "rfsc3", "rfsc4", "rfsc5", "rfsc6", "rfsc7", "rfsc8", "rfsc9", "rfsc10", "rfsc11", "rfsc12", "rfsc13", "rfsc14", "rfsc15", "rfsc16", "rfsc17", "rfsc18", "rfsc19")

c_rfsc_means_table <-
   # use admin data set
  ftp_ar %>% 
  # filter for control group
  filter(e == 0) %>% 
  # isolate afdc/tanf payment columns
  select (all_of(rfsc_cols)) %>% 
  # calculate the mean of each column
  summarise(across(everything(), ~ mean (., na.rm = TRUE))) %>% 
  # convert means to percentages (multiply by 100)
  mutate(across(everything(), ~ . * 100)) %>% 
  # rename quarters for readability
  setNames(c("Quarter of RA*", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8", "Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18")) %>% 
  # transpose for readability
  pivot_longer(., cols = everything()) %>% 
  setNames(c("Quarter", "Any Food Stamp Receipt (%)"))

# print table
kable(c_rfsc_means_table, 
      # add a title
      caption = "Average Quarterly Food Stamp Receipt Rates among Control Group",
      # round to tenths place to mimic original table
      digits = 1) %>% 
  # set style
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  # add footnote to define "RA"
  footnote(symbol = "RA = Random Assignment")

```

*Are there any discrepancies?*

There is one major discrepancy between the "Average Quarterly Food Stamp
Receipt Rates among Control Group" table and "AR Table 1_U". While the
above reports that "36.7"% of the control group received food stamps in
Quarter 17, the original table reports "0.4". While 36.7 is similar to
the numbers reported in other quarters and follows the general trend,
0.4 is drastically different, implying some error in the original table.

There are a few minor discrepancies between the above tables and those
included in "AR Table 1_U" which are potentially due to differences in
rounding standards. For example, in "Average Quarterly Employment Rates
among Control Group" table, the mean employment rate of Quarter 3 is
39.6% (rounded from 39.645%); in the original table, this is reported as
39.7%. These discrepancies are never greater than 0.01%, and thus are
likely insignificant.

*While you are at it, comment on the levels of employment among the
people in the study population.*

On average, the control group appears to increase employment rates over
time.

```{r}
quarters <- c("Quarter of RA*" = "Quarter 0", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8", "Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18")
num_quarters <- as.numeric(gsub("Quarter ", "", quarters))

c_empq_means_table %>% 
  mutate(Quarter = gsub("of RA\\*", "0", Quarter)) %>% 
  mutate(quarter_numeric = as.numeric(gsub("Quarter ", "", Quarter))) %>% 
  ggplot(aes(x = quarter_numeric, y = `Employment Rate (%)`)) + 
  geom_point() + 
 geom_smooth(method = 'lm', formula = y~x) +
  labs(x = "Quarter after Random Assignment", title = "Control Group Employment Rate over Time")

```

(4) *Compare your results to those in Table B.1 of Bloom et al (2000).
    Why are they different? How do you expect these differences to
    affect your results?*

The results in Table B.1 are slightly different (percentage rates in the
Bloom table are between 0.2 higher and 0.5 lower than those in the
results above).

As explained on Page 3 of the data documentation, the files have been
changed from their original form to maintain participant
confidentiality. This includes top coding some variables, rounding all
administrative record variables, and completely dropping some variables
(including most date variables). Bloom's analysis may have thus been
able to employ some of the dropped variables to make more precise
estimates, while ours did not include the same adjustments. Following up
on the the prior question, there may also be differing rounding
parameters utilized in the original tables.

We are told, however, that extensive testing has been conducted to
confirm that the results in Bloom are very close to those we created.
Thus, while we would expect the minor differences we experienced due to
some slight data discrepancies, we should not expect major
dissimilarities.

(5) *Test for balance. To do so, use all variables whose labels contain
    the string “cova:”. These are pre-RA variables that MDRC sometimes
    used as controls.*

```{r cova_table}
# create a table including only covariate columns and treatment dummy
# covariate columns can be identified by "cova" in column label

# https://stackoverflow.com/questions/67664582/is-it-possible-to-select-columns-based-on-variable-labels 

cova_table <- ftp_ar %>% 
  # select treatment dummy column 
  select("e", 
    # and columns where...
    where(~ {
    # the column's label
    column_label <- attributes(.)$label
    # is not null AND contains 'cova' (returns TRUE)
    !is.null(column_label) && str_detect(column_label, 'cova')
    }))
```

*Conduct two simple tests. First, regress each variable on the treatment
dummy. Comment on your choice of which standard errors to compute.*

```{r regress each, results = 'asis'}
# regress each covariate on the treatment dummy

# create table output
cova_regress <- cova_table %>% 
  # pivot the data to allow for grouping by covariate
  pivot_longer(cols = -e, names_to = "covariate", values_to = "value") %>% 
  group_by(covariate) %>% 
  # apply regression across each covariate
  group_map(~{
    # regress each covariate on treatment dummy
    regress <- lm(value ~ e, data = .)
    
    # calculate robust standard errors
    robust_se <- sqrt(diag(vcovHC(regress, type = "HC1")))
    
    # tidy the regression table
    regress_table <- tidy(regress) %>% 
      # including covariates as strings
      mutate(covariate = toString(.y)) %>% 
      # and adding in robust_se
      mutate(robust_se = robust_se) %>% 
      # removing other standard error
      select(!std.error)
    
    regress_table
  }) %>% 
  bind_rows() %>% 
  ungroup()

# print a pretty table
data.frame(cova_regress) %>%
  # filtering for term
  filter(term == "e") %>% 
  # arrange columns
  select(term, covariate, estimate, robust_se, statistic, p.value) %>% 
  # set column names
  setNames(c("Term (Treatment Status)", "Covariate", "Regression Coefficient Estimate", "Robust Standard Error", "T-Statistic", "P-Value")) %>%
kable(digits = 3, caption = "Regression Results by Covariate") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

This test computes robust standard errors to adjust for
heteroskedasticity. Included covariates, such as earnings and
employment, may have greater variance among certain groups in the
population (i.e., they may be heteroskedastic). To adjust for
heteroskedasticity, we should calculate robust standard errors, which
will allow for better modeling.

*How many times do you reject the null at the 5% level?*

```{r rejecting null}
# identify n rows where p value <0.05 (reject null)
cova_regress %>% 
  filter(term == "e") %>% 
   filter(p.value <0.05) %>% 
  nrow(.)

```

*Roughly how many rejections would you expect under the null of random
assignment?*

We reject the null 0 times, as the statistical significance of each
covariate regression never exceeds 0.05. Under the null of random
assignment, we do not expect statistically significant correlation
between the treatment dummy and our covariates. Random assignment should
purportedly evenly distribute the characteristics between the treatment
and control groups.

*Second, regress the treatment dummy on all the variables together and
conduct a joint test for balance.*

```{r regress all}
# regress treatment dummy on all variables together
allcova_regress <- lm(e ~ ., data = cova_table)
# summary(allcova_regress)
# conduct joint test for balance
# per Ed Discussion board, installing car

# cor(cova_table$yrearn, cova_table$yrearnsq)
# excluding yrearnsq

cova_balance <- linearHypothesis(allcova_regress, c("yr2adc = 0", "yradc = 0", "yr2rec = 0", "yrrec = 0", "yrkrec = 0", "yrearn = 0", "yr2earn = 0", "yremp = 0", "yr2emp = 0", "yrkemp = 0", "yr2kemp = 0", "yr2fs = 0", "yrfs = 0", "yr2rfs = 0", "yrrfs = 0", "yrkrfs = 0"))

cova_balance %>% 
  kable(digits = 3, caption = "Joint Hypothesis Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

*Explain your choice of degrees of freedom in computing this test.*

This test only includes 16 degrees of freedom despite having 17
covariates included in the model. This is because variable "yrearnsq",
as the square of "yrearn", is too closely correlated with "yrearn" to
properly model. This can be seen by running
"cor(cova_table$yrearn, cova_table$yrearnsq)", which reflects a very
high (0.875) correlation coefficient.

*Do the two tests yield the same conclusion? What do you conclude about
the overall balance of the sample?*

The second test returned F-Value is 0.83, and it's correlated P-Value is
0.65 (much greater than 0.05). Therefore, we fail to reject the null
hypothesis that treatment assignment is unrelated to the covariates.
This yields a similar conclusion to the first test, which suggests that
there are no systematic differences between the average covariate values
between the treatment and control group.

Because both tests fail to reject the null hypothesis that treatment
assignment is unrelated to the covariates, we can conclude that
treatment assignment in this study was in fact random. The sample is
thus balanced, with respect to the covariates.

(6) *Returning to the post-RA data, estimate simple treatment effects
    such as those shown in “Annotated AR table 1_U.xls.” Can you
    replicate those results? Restrict attention again to the variables
    from question (3).*


(a) *quarterly employment, quarter of RA to quarter 18;*

```{r treatment qemp mean}
# create vector of employment quarter columns 1-19

# for quarter of RA to quarter 18, this will be quarter 1 to 19 in data set
# see data documentation pg. 19 for more info
empq_cols <- c("empq1", "empq2", "empq3", "empq4", "empq5", "empq6", "empq7", "empq8", "empq9", "empq10", "empq11", "empq12", "empq13", "empq14", "empq15", "empq16", "empq17", "empq18", "empq19")

t_empq_means_table <- 
  # use admin data set
  ftp_ar %>% 
  # filter for treatment group
  filter(e == 1) %>% 
  # isolate employment quarter columns
  select (all_of(empq_cols)) %>% 
  # calculate the mean of each column
  summarise(across(everything(), ~ mean (., na.rm = TRUE))) %>% 
  # convert means to percentages (multiply by 100)
  mutate(across(everything(), ~ . * 100)) %>% 
  # rename quarters for readability
  setNames(c("Quarter of RA*", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8","Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18" )) %>% 
  # transpose for readability
  pivot_longer(., cols = everything()) %>% 
  setNames(c("Quarter", "Treatment Employment Rate (%)"))

# print comparison table
t_empq_means_table %>% 
  mutate("Control Employment Rate (%)" = c_empq_means_table$`Employment Rate (%)`) %>% 
  mutate("Difference" = .$`Treatment Employment Rate (%)` - .$'Control Employment Rate (%)') %>% 
kable(.,  
      # add a title
      caption = "Average Quarterly Employment Rates",
      # round to tenths place to mimic original table
      digits = 1) %>% 
  # set style
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  # add footnote to define "RA"
  footnote(symbol = "RA = Random Assignment")
```
(b) *any AFDC/TANF payments, quarter of RA to quarter 18;*

```{r treatment any afdc mean}
any_afdc_cols <- c("recc1", "recc2", "recc3", "recc4", "recc5", "recc6", "recc7", "recc8", "recc9", "recc10", "recc11", "recc12", "recc13", "recc14", "recc15", "recc16", "recc17", "recc18", "recc19")

t_afdc_means_table <-
  # use admin data set
  ftp_ar %>% 
  # filter for treatment group
  filter(e == 1) %>% 
  # isolate afdc/tanf payment columns
  select (all_of(any_afdc_cols)) %>% 
  # calculate the mean of each column
  summarise(across(everything(), ~ mean (., na.rm = TRUE))) %>% 
  # convert means to percentages (multiply by 100)
  mutate(across(everything(), ~ . * 100)) %>% 
  # rename quarters for readability
  setNames(c("Quarter of RA*", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8", "Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18")) %>% 
  # transpose for readability
  pivot_longer(., cols = everything()) %>% 
  setNames(c("Quarter", "Treatment AFDC/TANF Receipt (%)"))

# print comparison table
t_afdc_means_table %>% 
  mutate("Control AFDC/TANF Rate (%)" = c_afdc_means_table$`Any AFDC/TANF Receipt (%)`) %>% 
  mutate("Difference" = .$`Treatment AFDC/TANF Receipt (%)` - .$'Control AFDC/TANF Rate (%)') %>% 
kable(.,  
      # add a title
      caption = "Average Quarterly AFDC/TANF Rates",
      # round to tenths place to mimic original table
      digits = 1) %>% 
  # set style
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  # add footnote to define "RA"
  footnote(symbol = "RA = Random Assignment")


```

(b) *any Food Stamp payments, quarter of RA to quarter 18;*

```{r treatment any food stamp mean}
rfsc_cols <- c("rfsc1", "rfsc2", "rfsc3", "rfsc4", "rfsc5", "rfsc6", "rfsc7", "rfsc8", "rfsc9", "rfsc10", "rfsc11", "rfsc12", "rfsc13", "rfsc14", "rfsc15", "rfsc16", "rfsc17", "rfsc18", "rfsc19")

t_rfsc_means_table <-
   # use admin data set
  ftp_ar %>% 
  # filter for treatment group
  filter(e == 1) %>% 
  # isolate afdc/tanf payment columns
  select (all_of(rfsc_cols)) %>% 
  # calculate the mean of each column
  summarise(across(everything(), ~ mean (., na.rm = TRUE))) %>% 
  # convert means to percentages (multiply by 100)
  mutate(across(everything(), ~ . * 100)) %>% 
  # rename quarters for readability
  setNames(c("Quarter of RA*", "Quarter 1", "Quarter 2", "Quarter 3", "Quarter 4", "Quarter 5", "Quarter 6", "Quarter 7", "Quarter 8", "Quarter 9", "Quarter 10", "Quarter 11", "Quarter 12", "Quarter 13", "Quarter 14", "Quarter 15", "Quarter 16", "Quarter 17", "Quarter 18")) %>% 
  # transpose for readability
  pivot_longer(., cols = everything()) %>% 
  setNames(c("Quarter", "Treatment Food Stamp Receipt (%)"))

# print comparison table
t_rfsc_means_table %>% 
  mutate("Control Food Stamp Receipt (%)" = c_rfsc_means_table$`Any Food Stamp Receipt (%)`) %>% 
  mutate("Difference" = .$`Treatment Food Stamp Receipt (%)` - .$'Control Food Stamp Receipt (%)') %>% 
kable(.,  
      # add a title
      caption = "Average Quarterly Food Stamp Receipt Rates",
      # round to tenths place to mimic original table
      digits = 1) %>% 
  # set style
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  # add footnote to define "RA"
  footnote(symbol = "RA = Random Assignment")

```


(7) What type of treatment effects do your estimates identify? How do
    you interpret them?

(8) To distinguish the effects of the time limits from the other
    components of the reform program, Grogger and Michalopoulos proposed
    to estimate the effects of FTP by the age of the youngest child in
    the family. You’ll do the same. This will require knowing who was
    assigned to a 24-month time limit, and who was assigned a 36- month
    time limit. Use the actual time limit for the treatment group and
    the imputed time limit for the control group. Find and tabulate the
    relevant variables. How many observations are in each group?

(9) Define four age groups as a function of the age of the youngest
    child in the family during the quarter of random assignment: (0)
    less than 3 years; (i) 3-8 years; (ii) 9- 14 years (for families
    with a 36-month time limit) or 9-15 years (for families with a
    24-month time limit); (iv) 15-19 (for families with a 36-month time
    limit) or 16-19 years (for families with a 24-month time limit). By
    treatment status, how many families are in each age group?

(10) Estimate age-group-specific treatment effects of FTP on number of
     months of AFDC receipt and Food Stamp receipt, and number of
     quarters of employment , during the first 24 months after random
     assignment. Are the estimates of the effects on AFDC receipt
     consistent with the hypothesis that time limits should have
     stronger effects, the younger the youngest child in the family?
     Note that families in age group (iv) (from Question 9) shouldn’t be
     affected by time limits at all. They are only affected by the other
     components of the program, so they provide an estimate of the
     collective effect of those components. Under some assumptions, you
     can estimate the effects of the time limit on the younger age
     groups as the difference between the age-group specific effects of
     the reform and the effect of the reform on the oldest age group.

(11) To your regression from (10) add a dummy that equals one if the
     family has a 36-month time limit. Reassess the evidence for whether
     time limits have stronger effects, the younger the youngest child
     in the family. Explain the difference between these estimates and
     those that do not control for the length of the time limit.
